---
title: "Regression Models Assignment"
author: "Sara Rafi"
output: pdf_document
---

##Executive Summary

The motot trance car roads data comprises of fuel compsumtion data along with 10 aspects of automobile design and performance. The sample consists of data for 32 autmobiles.

This report attempts to answer the following questions:

1. Is an automatic or manual transmission better for MPG

2. Quantify the MPG difference between automatic and manual transmissions

Preliminary data exploration shows an increase in MPG for autmobiles with automotic transmission. However, a regression analysis fails to provide evidence of such a relation.

##Data 

The dataframe comroses fuel consumption and ten aspects of automobile design performace for thirtytwo automobiles. The dataframes consists of the the following variables.

```{r}
names(mtcars)
```

The characteristics of these variables are:

```{r}
str(mtcars)

```
###Coverting variables to factors

```{r}
mtcars$am     <- as.factor(mtcars$am)
mtcars$cyl    <- as.factor(mtcars$cyl)
mtcars$vs     <- as.factor(mtcars$vs)
mtcars$gear   <- as.factor(mtcars$gear)

```
###Exploratory Analysis

As shown in Figure 1 in the Appendix, plotting the MPG for the manual vs the automatic transmission cars shows that there is a difference in the MPG for the teo transmissions type. The  with automatic models have a higher MPG when compared to automatic cars.

##Model selection

The model selection is based on the Akaine Information Criteria(AIC). AIC selects a models a model by estimating and comparing the quality of each model in comparison to other models. The AIC value of a model is calculated by the followinf formula:

$$AIC=2k- 2ln(L)$$

In this formula *k is the number of estimated paramters in the model and *L is the maximised value of the likelihood function for the model. The preferred model is the one with the minimum AIC value.

This method takes in account for the possibility if interaction between different parameters.

For our analysis the AIC stages are as follows:
Our final model, with the lowes AIC is:

```{r}
fit1 <- lm(mpg ~ 1, data=mtcars)  # minimum possible regression model
fit <- step(fit1, direction="both", #adding parameters stepwise
            scope=(~ am:(cyl+disp+hp+drat+wt+qsec+vs+gear+carb))) 
```

```{r}
summary(fit)
```
The p-values also indicate the signicance of the variables. 

##Interpretation of Model

The final model can be written as:

  $$Y=\beta_0+\beta_1x_{weight}+1(am)(\beta_2-\beta_1)x_{weight}+\beta_3x_{qsec}+1(am)(\beta_4-\beta_3)x_{qsec}+e$$

where: am= 1 for automatic transmission
       am= 0 for manual transmission

The coefficients corresponding to the smmary table above are:

$\beta_0=intercept$

$\beta_1=am0:wt$ 

$\beta_2=am1:wt$  

$\beta_3=am0:qsec$ 

$\beta_4=am1:qsec$ 

##Residual Analysis 

###Graphical residual analysis

We assume that the residuals have constant variance and follow a normal distrubution. From Figure 2(see Appendix), we observer the following:

1. The residuals are scattered randomly around zero.

2. The residuals roughly form a "horizontal band" around the 0 line. This suggests that the variances of the error terms are equal.

3. The quantiles for the residuals fall close to the theoretical normal quantiles in the Q-Q plot. 

##Conclusion

Comparing the model for automatic and manual transmission we get: 

$$(-2.9233)x_{weight}+0.6126x_{qsec}$$

The above expression summarises the difference between automatic and manual transmission.  

In order for automatic to be better than manual the expression would have a value greater than 0. Thus, we can write it as following:

$$(-2.9233)x_{weight}+0.6126x_{qsec}>0$$                    

```{r}
library(plyr)

x<-ddply(mtcars, .(am), summarize,  wt_mean=mean(wt), qsec_mean=mean(qsec))

test_am<-(-2.9233)*x[1,2]+0.6126*x[1,3]
test_mu<-(-2.9233)*x[2,2]+0.6126*x[2,3]
test_am
test_mu
```
When we use our data to test the above expression, we get a positive result for both munual and automatic transmission. Thus, we are unable to find evidence that automatic cars are better for MPG than manual.

##Appendix

###Figure 1
```{r}
boxplot(mpg~am,data=mtcars,main="Averages",varwidth=TRUE, col=c("green","blue"),
        names=c("Manual","Automatic")) 
```
###Figure 2
```{r}
par(mfrow = c(1,2),mar = c(2, 2, 3, 2))
plot(fit, which=1)
plot(fit, which=2)
```
